---
title: "Statistical Tests for Categorical Outcomes"
author: "Stats Tests Lecture 2025"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
```

# Statistical Tests for Categorical Outcomes

This document covers statistical tests for categorical outcomes, focusing on tests of association and independence between categorical variables.

## 1. Chi-Square Test of Independence

The Chi-Square test of independence examines whether two categorical variables are associated or independent.

**Assumptions:**
- Independence of observations
- Expected frequency count ≥ 5 in at least 80% of cells
- Expected frequency count > 1 in all cells
- Random sampling

**Example:**
```{r chi-square-test}
# Generate sample data: relationship between smoking status and lung disease
set.seed(111)
n <- 200

# Create a contingency table
smoking <- sample(c("Smoker", "Non-smoker"), n, replace = TRUE, prob = c(0.3, 0.7))
lung_disease <- character(n)

# Simulate association between smoking and lung disease
for (i in 1:n) {
  if (smoking[i] == "Smoker") {
    lung_disease[i] <- sample(c("Disease", "No Disease"), 1, prob = c(0.4, 0.6))
  } else {
    lung_disease[i] <- sample(c("Disease", "No Disease"), 1, prob = c(0.15, 0.85))
  }
}

# Create data frame
disease_data <- data.frame(
  smoking = factor(smoking),
  lung_disease = factor(lung_disease)
)

# Create contingency table
cont_table <- table(disease_data$smoking, disease_data$lung_disease)
print("Contingency Table:")
print(cont_table)

# Add margins for better visualization
print("\nContingency Table with Margins:")
addmargins(cont_table) %>% kable()

# Visualize the data
disease_summary <- disease_data %>%
  group_by(smoking, lung_disease) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(smoking) %>%
  mutate(proportion = count / sum(count))

ggplot(disease_summary, aes(x = smoking, y = count, fill = lung_disease)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Lung Disease by Smoking Status",
       x = "Smoking Status", y = "Count",
       fill = "Lung Disease")

# Perform Chi-Square test
chi_square_result <- chisq.test(cont_table)
print("\nChi-Square Test Results:")
print(chi_square_result)

# Display expected frequencies
print("\nExpected Frequencies:")
print(chi_square_result$expected)

# Calculate effect size (Cramér's V)
n_total <- sum(cont_table)
chi_sq_stat <- chi_square_result$statistic
min_dim <- min(nrow(cont_table), ncol(cont_table)) - 1
cramers_v <- sqrt(chi_sq_stat / (n_total * min_dim))

# Interpretation
cat("\n=== INTERPRETATION ===\n")
cat("Chi-square statistic:", round(chi_square_result$statistic, 3), "\n")
cat("Degrees of freedom:", chi_square_result$parameter, "\n")
cat("P-value:", format(chi_square_result$p.value, scientific = FALSE), "\n")
cat("Cramér's V (effect size):", round(cramers_v, 3), "\n")

if (chi_square_result$p.value < 0.05) {
  cat("\nConclusion: There is a significant association between smoking status and lung disease (p < 0.05)\n")
} else {
  cat("\nConclusion: There is no significant association between smoking status and lung disease (p >= 0.05)\n")
}

# Effect size interpretation
cat("\nEffect Size Interpretation (Cramér's V):\n")
if (cramers_v < 0.1) {
  cat("- Small effect\n")
} else if (cramers_v < 0.3) {
  cat("- Medium effect\n")
} else {
  cat("- Large effect\n")
}
```

## 2. Fisher's Exact Test

Fisher's Exact Test is used when sample sizes are small and Chi-Square test assumptions are violated. It calculates the exact probability of observing the data.

**When to use:**
- Small sample sizes
- Expected cell frequencies < 5
- 2x2 contingency tables (can be extended to larger tables)

**Example 1: Small Sample Size**
```{r fisher-exact-test-small}
# Generate small sample data: treatment effect in a clinical trial
set.seed(222)

# Create a small 2x2 contingency table
treatment_table <- matrix(c(8, 2,   # Treatment group: 8 improved, 2 not improved
                           3, 7),  # Control group: 3 improved, 7 not improved
                         nrow = 2,
                         dimnames = list(Group = c("Treatment", "Control"),
                                       Outcome = c("Improved", "Not Improved")))

print("Contingency Table (Small Sample):")
print(treatment_table)
addmargins(treatment_table) %>% kable()

# Visualize the data
treatment_df <- as.data.frame(as.table(treatment_table))
names(treatment_df) <- c("Group", "Outcome", "Count")

ggplot(treatment_df, aes(x = Group, y = Count, fill = Outcome)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Treatment Outcomes (Small Sample)",
       x = "Group", y = "Count")

# Perform Fisher's Exact Test
fisher_result <- fisher.test(treatment_table)
print("\nFisher's Exact Test Results:")
print(fisher_result)

# Interpretation
cat("\n=== INTERPRETATION ===\n")
cat("P-value:", format(fisher_result$p.value, scientific = FALSE), "\n")
cat("Odds Ratio:", round(fisher_result$estimate, 3), "\n")
cat("95% Confidence Interval:", round(fisher_result$conf.int[1], 3), 
    "to", round(fisher_result$conf.int[2], 3), "\n")

if (fisher_result$p.value < 0.05) {
  cat("\nConclusion: There is a significant association between treatment and outcome (p < 0.05)\n")
} else {
  cat("\nConclusion: There is no significant association between treatment and outcome (p >= 0.05)\n")
}

# Odds ratio interpretation
if (fisher_result$estimate > 1) {
  cat("\nThe odds of improvement are", round(fisher_result$estimate, 2), 
      "times higher in the treatment group compared to control.\n")
} else {
  cat("\nThe odds of improvement are lower in the treatment group compared to control.\n")
}
```

**Example 2: Rare Event**
```{r fisher-exact-test-rare}
# Generate data for a rare adverse event study
set.seed(333)

# Create a 2x2 contingency table for rare event
adverse_event_table <- matrix(c(1, 49,   # New drug: 1 adverse event, 49 no event
                               5, 45),  # Standard drug: 5 adverse events, 45 no event
                             nrow = 2,
                             dimnames = list(Drug = c("New", "Standard"),
                                           Event = c("Adverse Event", "No Event")))

print("Contingency Table (Rare Event):")
print(adverse_event_table)
addmargins(adverse_event_table) %>% kable()

# Check Chi-Square assumptions
chi_check <- chisq.test(adverse_event_table)
print("\nExpected frequencies (Chi-Square assumptions check):")
print(chi_check$expected)
cat("\nNote: Expected frequencies < 5 detected, Fisher's Exact Test is appropriate.\n")

# Perform Fisher's Exact Test
fisher_rare_result <- fisher.test(adverse_event_table)
print("\nFisher's Exact Test Results:")
print(fisher_rare_result)

# Visualize
adverse_df <- as.data.frame(as.table(adverse_event_table))
names(adverse_df) <- c("Drug", "Event", "Count")

ggplot(adverse_df, aes(x = Drug, y = Count, fill = Event)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(title = "Adverse Events by Drug Type",
       x = "Drug", y = "Count") +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5))

# Interpretation
cat("\n=== INTERPRETATION ===\n")
cat("P-value:", format(fisher_rare_result$p.value, scientific = FALSE), "\n")
cat("Odds Ratio:", round(fisher_rare_result$estimate, 3), "\n")
cat("95% Confidence Interval:", round(fisher_rare_result$conf.int[1], 3), 
    "to", round(fisher_rare_result$conf.int[2], 3), "\n")

if (fisher_rare_result$p.value < 0.05) {
  cat("\nConclusion: There is a significant difference in adverse event rates between drugs (p < 0.05)\n")
} else {
  cat("\nConclusion: There is no significant difference in adverse event rates between drugs (p >= 0.05)\n")
}
```

## Comparison: Chi-Square vs Fisher's Exact Test

```{r comparison-table}
comparison_df <- data.frame(
  Aspect = c("Sample Size", "Expected Frequencies", 
             "Computation", "Accuracy", "Use Case"),
  `Chi-Square Test` = c("Large (n > 20-40)", 
                        "≥5 in 80% of cells",
                        "Asymptotic approximation",
                        "Approximate p-value",
                        "General contingency tables"),
  `Fisher's Exact Test` = c("Any size, especially small",
                            "No restrictions",
                            "Exact calculation",
                            "Exact p-value",
                            "Small samples, rare events")
)

kable(comparison_df, col.names = c("Aspect", "Chi-Square Test", "Fisher's Exact Test"),
      caption = "Comparison of Chi-Square and Fisher's Exact Tests")
```

## Decision Tree for Test Selection

```{r decision-tree, echo=FALSE, fig.height=4}
# Create a simple decision flowchart using text
cat("
╔═══════════════════════════════════════╗
║   Testing Association Between Two     ║
║      Categorical Variables            ║
╚═══════════════════════════════════════╝
                  │
                  ▼
        ┌─────────────────────┐
        │  Check Sample Size  │
        └─────────────────────┘
                  │
         ┌────────┴────────┐
         ▼                 ▼
    Large Sample      Small Sample
    (n > 40)         (n ≤ 40)
         │                 │
         ▼                 ▼
  ┌──────────────┐   ┌──────────────┐
  │ Check Exp.   │   │   Fisher's   │
  │ Frequencies  │   │  Exact Test  │
  └──────────────┘   └──────────────┘
         │
    ┌────┴────┐
    ▼         ▼
  All ≥ 5   Any < 5
    │         │
    ▼         ▼
Chi-Square  Fisher's
   Test    Exact Test
")
```

## Summary

This document covered two essential statistical tests for categorical outcomes:

1. **Chi-Square Test of Independence**
   - Tests association between two categorical variables
   - Requires adequate sample size and expected frequencies
   - Provides approximate p-values
   - Appropriate for large samples

2. **Fisher's Exact Test**
   - Alternative when Chi-Square assumptions are violated
   - Works with any sample size
   - Provides exact p-values
   - Especially useful for small samples and rare events

### Key Takeaways:

- Always check sample size and expected frequencies before choosing a test
- Fisher's Exact Test is more conservative but provides exact p-values
- Chi-Square Test is computationally simpler for large samples
- Visualize your data with contingency tables and bar plots
- Report effect sizes (e.g., Cramér's V, Odds Ratio) along with p-values
- Consider the context and practical significance of your findings

### When to Use Each Test:

**Use Chi-Square Test when:**
- Sample size is large (n > 40)
- Expected frequency ≥ 5 in at least 80% of cells
- All expected frequencies > 1

**Use Fisher's Exact Test when:**
- Sample size is small (n ≤ 40)
- Any expected frequency < 5
- Analyzing rare events
- You need exact p-values
