---
title: "Statistical Tests for Continuous Outcomes"
author: "Stats Tests Lecture 2025"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Statistical Tests for Continuous Outcomes

This document covers statistical tests for continuous outcomes, including parametric and non-parametric tests for comparing means.

## 1. Independent Samples t-Test

The independent samples t-test compares the means of two independent groups.

**Assumptions:**
- Independence of observations
- Normal distribution in each group
- Homogeneity of variance (equal variances)

**Example:**
```{r independent-t-test}
# Generate sample data: heights of males vs females
set.seed(123)
male_heights <- rnorm(30, mean = 175, sd = 10)
female_heights <- rnorm(30, mean = 165, sd = 10)

# Combine into a data frame
height_data <- data.frame(
  height = c(male_heights, female_heights),
  gender = factor(rep(c("Male", "Female"), each = 30))
)

# Visualize the data
ggplot(height_data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Height Distribution by Gender",
       x = "Gender", y = "Height (cm)")

# Perform independent samples t-test
t_test_result <- t.test(height ~ gender, data = height_data, var.equal = TRUE)
print(t_test_result)

# Interpretation
cat("\nInterpretation:\n")
cat("The p-value is:", format(t_test_result$p.value, scientific = FALSE), "\n")
if (t_test_result$p.value < 0.05) {
  cat("There is a significant difference between the two groups (p < 0.05)\n")
} else {
  cat("There is no significant difference between the two groups (p >= 0.05)\n")
}
```

## 2. Paired Samples t-Test

The paired samples t-test compares the means of two related groups (e.g., before and after measurements).

**Assumptions:**
- Paired observations
- Normal distribution of differences
- Independence of pairs

**Example:**
```{r paired-t-test}
# Generate sample data: weight before and after a diet program
set.seed(456)
n <- 25
weight_before <- rnorm(n, mean = 80, sd = 12)
weight_after <- weight_before - rnorm(n, mean = 3, sd = 2)  # Average loss of 3 kg

# Create data frame
weight_data <- data.frame(
  participant = 1:n,
  before = weight_before,
  after = weight_after
)

# Visualize the data
weight_long <- weight_data %>%
  pivot_longer(cols = c(before, after), names_to = "time", values_to = "weight")

ggplot(weight_long, aes(x = time, y = weight, group = participant)) +
  geom_line(color = "gray") +
  geom_point(aes(color = time), size = 3) +
  theme_minimal() +
  labs(title = "Weight Before and After Diet Program",
       x = "Time", y = "Weight (kg)")

# Perform paired samples t-test
paired_t_result <- t.test(weight_data$before, weight_data$after, paired = TRUE)
print(paired_t_result)

# Interpretation
cat("\nInterpretation:\n")
cat("Mean difference:", round(paired_t_result$estimate, 2), "kg\n")
cat("The p-value is:", format(paired_t_result$p.value, scientific = FALSE), "\n")
if (paired_t_result$p.value < 0.05) {
  cat("There is a significant difference between before and after (p < 0.05)\n")
} else {
  cat("There is no significant difference between before and after (p >= 0.05)\n")
}
```

## 3. One-Sample t-Test

The one-sample t-test compares the mean of a sample to a known or hypothesized population mean.

**Assumptions:**
- Independence of observations
- Normal distribution
- Random sampling

**Example:**
```{r one-sample-t-test}
# Generate sample data: IQ scores (population mean is 100)
set.seed(789)
sample_iq <- rnorm(40, mean = 105, sd = 15)

# Visualize the data
ggplot(data.frame(iq = sample_iq), aes(x = iq)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 100, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = 100, y = 6, label = "Population Mean = 100", 
           color = "red", hjust = -0.1) +
  theme_minimal() +
  labs(title = "Sample IQ Scores Distribution",
       x = "IQ Score", y = "Frequency")

# Perform one-sample t-test (testing against population mean of 100)
one_sample_result <- t.test(sample_iq, mu = 100)
print(one_sample_result)

# Interpretation
cat("\nInterpretation:\n")
cat("Sample mean:", round(mean(sample_iq), 2), "\n")
cat("Hypothesized population mean:", 100, "\n")
cat("The p-value is:", format(one_sample_result$p.value, scientific = FALSE), "\n")
if (one_sample_result$p.value < 0.05) {
  cat("The sample mean is significantly different from 100 (p < 0.05)\n")
} else {
  cat("The sample mean is not significantly different from 100 (p >= 0.05)\n")
}
```

## 4. Mann-Whitney U Test (Wilcoxon Rank-Sum Test)

The Mann-Whitney U test is a non-parametric alternative to the independent samples t-test. It compares the distributions of two independent groups without assuming normality.

**When to use:**
- Data is not normally distributed
- Ordinal data
- Small sample sizes

**Example:**
```{r mann-whitney-test}
# Generate skewed sample data: reaction times in two groups
set.seed(321)
group_a <- rexp(25, rate = 0.5)  # Exponentially distributed
group_b <- rexp(25, rate = 0.7)

reaction_data <- data.frame(
  time = c(group_a, group_b),
  group = factor(rep(c("Group A", "Group B"), each = 25))
)

# Visualize the data
ggplot(reaction_data, aes(x = group, y = time, fill = group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Reaction Time Distribution by Group",
       x = "Group", y = "Reaction Time (seconds)")

# Perform Mann-Whitney U test
mann_whitney_result <- wilcox.test(time ~ group, data = reaction_data)
print(mann_whitney_result)

# Interpretation
cat("\nInterpretation:\n")
cat("Median Group A:", round(median(group_a), 2), "\n")
cat("Median Group B:", round(median(group_b), 2), "\n")
cat("The p-value is:", format(mann_whitney_result$p.value, scientific = FALSE), "\n")
if (mann_whitney_result$p.value < 0.05) {
  cat("There is a significant difference between the two groups (p < 0.05)\n")
} else {
  cat("There is no significant difference between the two groups (p >= 0.05)\n")
}
```

## 5. Wilcoxon Signed-Rank Test

The Wilcoxon signed-rank test is a non-parametric alternative to the paired samples t-test. It compares two related samples without assuming normality.

**When to use:**
- Paired data is not normally distributed
- Ordinal paired data
- Small sample sizes

**Example:**
```{r wilcoxon-signed-rank-test}
# Generate paired sample data: pain scores before and after treatment
set.seed(654)
n <- 20
pain_before <- sample(5:10, n, replace = TRUE)  # Pain scale 1-10
pain_after <- pmax(1, pain_before - sample(0:4, n, replace = TRUE))  # Reduction

pain_paired_data <- data.frame(
  patient = 1:n,
  before = pain_before,
  after = pain_after
)

# Visualize the data
pain_long <- pain_paired_data %>%
  pivot_longer(cols = c(before, after), names_to = "time", values_to = "pain_score")

ggplot(pain_long, aes(x = time, y = pain_score, group = patient)) +
  geom_line(color = "gray") +
  geom_point(aes(color = time), size = 3) +
  theme_minimal() +
  labs(title = "Pain Scores Before and After Treatment",
       x = "Time", y = "Pain Score (1-10)")

# Perform Wilcoxon signed-rank test
wilcoxon_result <- wilcox.test(pain_paired_data$before, 
                                pain_paired_data$after, 
                                paired = TRUE)
print(wilcoxon_result)

# Interpretation
cat("\nInterpretation:\n")
cat("Median pain before:", median(pain_before), "\n")
cat("Median pain after:", median(pain_after), "\n")
cat("The p-value is:", format(wilcoxon_result$p.value, scientific = FALSE), "\n")
if (wilcoxon_result$p.value < 0.05) {
  cat("There is a significant difference between before and after treatment (p < 0.05)\n")
} else {
  cat("There is no significant difference between before and after treatment (p >= 0.05)\n")
}
```

## Summary

This document covered five essential statistical tests for continuous outcomes:

1. **Independent Samples t-Test**: Compares means of two independent groups (parametric)
2. **Paired Samples t-Test**: Compares means of two related groups (parametric)
3. **One-Sample t-Test**: Compares sample mean to a known population mean (parametric)
4. **Mann-Whitney U Test**: Non-parametric alternative to independent samples t-test
5. **Wilcoxon Signed-Rank Test**: Non-parametric alternative to paired samples t-test

### Key Takeaways:

- Use parametric tests (t-tests) when data is normally distributed
- Use non-parametric tests when data is skewed or ordinal
- Always visualize your data before choosing a test
- Check assumptions before applying parametric tests
- Interpret results in context with effect sizes and practical significance
