---
title: "Hypothesis testing. Classical statistical tests"
author: "Jelena Čuklina & Mariia Chernigovsaya"
date: "October 2025"
output:
  html_document: default
---
# Setup


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-packages}
library(dplyr)      
library(ggplot2)
```

```{r load-data}
# Load the NHANES dataset
```

Prepare a clean dataset for analysis
```{r prepare-data}
# Focus on key variables for the exercise
nhanes_clean <- NHANES %>%
  select(Gender, Age, BMI, Diabetes, BPSysAve, BPDiaAve) %>%
  # Remove rows with any missing data for simplicity in these exercises
  na.omit() %>%
  # Filter for Age >= 18 for adult studies
  filter(Age >= 18)

# Check the structure of the data for students
# str(nhanes_clean)
```


# Tests for continuous data
## Parametric tests

### z-test

The scenario:Known Population Mean ($\mu_0$): Based on historical data, the mean BMI for adults in the U.S. is $\mathbf{28.5}$.Study Question: Does the mean BMI of a specific subsample (older adults, age $\ge 60$) significantly differ from the general historical population mean of $28.5$?

```{r}
# Example of one-sample z-test
# Define the null hypothesis mean (mu_0)
mu_0_sys <- 120

# Perform the one-sample t-test
t_result_one_sample <- t.test(nhanes_clean$BPSysAve, mu = mu_0_sys)

cat("\n--- Results for One-Sample t-test (BPSysAve vs. 120) ---\n")
print(t_result_one_sample)

# Question for Clinicians:
# How does the p-value and the 95% Confidence Interval inform you about
# the general SBP status of this population relative to the 120 threshold?
```

### t-test

We will cover three types of t-tests:
  * one-sample t-test
  * independent two-sample t-test
  * paired t-test

## Non-parametric tests
### Wilcoxon signed-rank test
### Mann-Whitney U test (Wilcoxon rank-sum test)

## Other tests for continuous data (to be covered in later lectures):
- ANOVA
- ANCOVA (TBC)
- Correlation (Pearson, Spearman)
- Linear regression

# Tests for categorical data
## Parametric tests
### Chi-squared test for goodness of fit
### Chi-squared test for independence
### Fisher's exact test

## Non-parametric tests
### McNemar's test
### Cochran's Q test