---
title: "Homework: Statistical Tests Practice"
author: "Stats Tests Lecture 2025"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Homework: Statistical Tests Practice

This homework assignment covers all the statistical tests learned in the lecture series. Complete each exercise by analyzing the provided data and answering the questions.

---

## Part 1: Continuous Outcomes

### Exercise 1.1: Independent Samples t-Test

**Scenario:** A pharmaceutical company wants to test whether a new sleep medication increases sleep duration. They randomly assign 30 patients to receive the medication and 30 patients to receive a placebo.

```{r exercise-1-1-data}
# Sleep duration data (in hours)
set.seed(2025)
sleep_medication <- rnorm(30, mean = 7.5, sd = 1.2)
sleep_placebo <- rnorm(30, mean = 6.8, sd = 1.3)

sleep_data <- data.frame(
  duration = c(sleep_medication, sleep_placebo),
  group = factor(rep(c("Medication", "Placebo"), each = 30))
)

# Display first few rows
head(sleep_data)
```

**Tasks:**
1. Create a boxplot to visualize the sleep duration by group
2. Perform an independent samples t-test
3. State the null and alternative hypotheses
4. Interpret the results (α = 0.05)
5. Calculate and interpret the 95% confidence interval for the difference in means

**Your code here:**
```{r exercise-1-1-solution}
# Write your solution here



```

---

### Exercise 1.2: Paired Samples t-Test

**Scenario:** A fitness coach wants to evaluate the effectiveness of a 6-week training program. They measure the 5K running time (in minutes) of 20 participants before and after the program.

```{r exercise-1-2-data}
# Running time data (in minutes)
set.seed(2026)
participants <- 20
time_before <- rnorm(participants, mean = 28, sd = 4)
time_after <- time_before - rnorm(participants, mean = 2, sd = 1.5)

running_data <- data.frame(
  participant_id = 1:participants,
  before_training = time_before,
  after_training = time_after
)

head(running_data)
```

**Tasks:**
1. Create a visualization showing the before and after measurements
2. Perform a paired samples t-test
3. Calculate the mean difference and its 95% confidence interval
4. Interpret whether the training program was effective (α = 0.05)
5. Discuss whether the difference is practically significant

**Your code here:**
```{r exercise-1-2-solution}
# Write your solution here



```

---

### Exercise 1.3: One-Sample t-Test

**Scenario:** A coffee shop claims that their espresso shots contain exactly 75mg of caffeine. A consumer watchdog group tests 25 random espresso shots to verify this claim.

```{r exercise-1-3-data}
# Caffeine content data (in mg)
set.seed(2027)
caffeine_content <- rnorm(25, mean = 78, sd = 5)

caffeine_data <- data.frame(
  shot_id = 1:25,
  caffeine_mg = caffeine_content
)

head(caffeine_data)
```

**Tasks:**
1. Create a histogram of the caffeine content with a reference line at 75mg
2. Perform a one-sample t-test against the claimed value of 75mg
3. State your conclusion about whether the claim is accurate (α = 0.05)
4. Calculate the sample mean and its 95% confidence interval

**Your code here:**
```{r exercise-1-3-solution}
# Write your solution here



```

---

### Exercise 1.4: Mann-Whitney U Test

**Scenario:** A researcher is studying stress levels (measured on a non-normal ordinal scale from 1-20) in two different work environments: open office vs. private office.

```{r exercise-1-4-data}
# Stress level data (ordinal scale 1-20, higher = more stress)
set.seed(2028)
stress_open <- sample(8:20, 28, replace = TRUE, prob = (8:20)^1.5)
stress_private <- sample(5:18, 28, replace = TRUE, prob = (5:18)^1.2)

stress_data <- data.frame(
  stress_level = c(stress_open, stress_private),
  office_type = factor(rep(c("Open Office", "Private Office"), each = 28))
)

head(stress_data)
```

**Tasks:**
1. Create a boxplot to visualize stress levels by office type
2. Explain why a non-parametric test is more appropriate than a t-test for this data
3. Perform a Mann-Whitney U test
4. Compare the median stress levels between the two groups
5. Interpret the results (α = 0.05)

**Your code here:**
```{r exercise-1-4-solution}
# Write your solution here



```

---

### Exercise 1.5: Wilcoxon Signed-Rank Test

**Scenario:** A pain management clinic uses a treatment protocol. They measure pain intensity (on a 0-10 scale) in 15 patients before and after treatment. The data is ordinal and not normally distributed.

```{r exercise-1-5-data}
# Pain intensity data (scale 0-10)
set.seed(2029)
pain_before <- sample(6:10, 15, replace = TRUE)
pain_after <- pmax(2, pain_before - sample(1:4, 15, replace = TRUE))

pain_data <- data.frame(
  patient_id = 1:15,
  pain_before = pain_before,
  pain_after = pain_after
)

head(pain_data)
```

**Tasks:**
1. Create a visualization showing pain scores before and after treatment
2. Explain why Wilcoxon signed-rank test is preferred over paired t-test
3. Perform a Wilcoxon signed-rank test
4. Compare the median pain scores
5. Interpret whether the treatment was effective (α = 0.05)

**Your code here:**
```{r exercise-1-5-solution}
# Write your solution here



```

---

## Part 2: Categorical Outcomes

### Exercise 2.1: Chi-Square Test of Independence

**Scenario:** A university wants to investigate whether there's an association between student major (STEM vs. Non-STEM) and participation in study abroad programs (Yes vs. No). They survey 200 students.

```{r exercise-2-1-data}
# Create survey data
set.seed(2030)
majors <- sample(c("STEM", "Non-STEM"), 200, replace = TRUE, prob = c(0.6, 0.4))
study_abroad <- character(200)

for (i in 1:200) {
  if (majors[i] == "STEM") {
    study_abroad[i] <- sample(c("Yes", "No"), 1, prob = c(0.25, 0.75))
  } else {
    study_abroad[i] <- sample(c("Yes", "No"), 1, prob = c(0.45, 0.55))
  }
}

student_data <- data.frame(
  major = factor(majors),
  study_abroad = factor(study_abroad)
)

# Display contingency table
table(student_data$major, student_data$study_abroad)
```

**Tasks:**
1. Create a contingency table with margins
2. Create a bar plot to visualize the relationship
3. Check whether Chi-Square test assumptions are met
4. Perform a Chi-Square test of independence
5. Calculate and interpret Cramér's V effect size
6. Write a conclusion about the relationship between major and study abroad participation

**Your code here:**
```{r exercise-2-1-solution}
# Write your solution here



```

---

### Exercise 2.2: Fisher's Exact Test

**Scenario:** A clinical trial tests a new treatment for a rare disease. Due to the rarity of the disease, only 30 patients are enrolled: 15 receive the new treatment and 15 receive standard care. The outcome is recovery (Yes/No).

```{r exercise-2-2-data}
# Create clinical trial data
set.seed(2031)
treatment_group <- c(rep("New Treatment", 15), rep("Standard Care", 15))
recovery <- c(
  sample(c("Recovered", "Not Recovered"), 15, replace = TRUE, prob = c(0.6, 0.4)),
  sample(c("Recovered", "Not Recovered"), 15, replace = TRUE, prob = c(0.35, 0.65))
)

trial_data <- data.frame(
  treatment = factor(treatment_group),
  outcome = factor(recovery)
)

# Display contingency table
trial_table <- table(trial_data$treatment, trial_data$outcome)
print(trial_table)
```

**Tasks:**
1. Create a visualization of the trial results
2. Explain why Fisher's Exact Test is more appropriate than Chi-Square test for this data
3. Perform Fisher's Exact Test
4. Interpret the odds ratio and its confidence interval
5. Write a conclusion about treatment effectiveness (α = 0.05)

**Your code here:**
```{r exercise-2-2-solution}
# Write your solution here



```

---

## Part 3: Comprehensive Analysis

### Exercise 3.1: Choosing the Right Test

For each scenario below, identify which statistical test is most appropriate and explain why:

**Scenario A:** Comparing average exam scores between students who attended review sessions (n=45) and those who didn't (n=52). Scores are normally distributed.

**Scenario B:** Comparing customer satisfaction ratings (scale 1-5) before and after a service improvement in the same 20 customers. Data is ordinal.

**Scenario C:** Testing whether the average height of a sample of 100 adult males differs from the national average of 175 cm.

**Scenario D:** Investigating the relationship between gender (Male/Female) and voting preference (Candidate A/Candidate B) in a sample of 300 voters.

**Scenario E:** Comparing recovery rates between two treatments in a study with only 8 patients in each group.

**Your answers here:**
```{r exercise-3-1-solution}
# Write your answers as comments:

# Scenario A:
# Test: 
# Reason: 

# Scenario B:
# Test: 
# Reason: 

# Scenario C:
# Test: 
# Reason: 

# Scenario D:
# Test: 
# Reason: 

# Scenario E:
# Test: 
# Reason: 

```

---

### Exercise 3.2: Real-World Application

**Scenario:** You are a data analyst at a healthcare organization. The hospital administrator asks you to analyze whether a new patient education program affects readmission rates. You have data from 150 patients before the program and 150 patients after the program implementation.

```{r exercise-3-2-data}
# Readmission data
set.seed(2032)
before_program <- sample(c("Readmitted", "Not Readmitted"), 150, 
                        replace = TRUE, prob = c(0.22, 0.78))
after_program <- sample(c("Readmitted", "Not Readmitted"), 150, 
                       replace = TRUE, prob = c(0.14, 0.86))

readmission_data <- data.frame(
  period = factor(rep(c("Before Program", "After Program"), each = 150)),
  outcome = factor(c(before_program, after_program))
)

table(readmission_data$period, readmission_data$outcome)
```

**Tasks:**
1. Identify the type of data (continuous or categorical)
2. Choose and justify the appropriate statistical test
3. Perform the test and interpret results
4. Calculate an appropriate effect size measure
5. Write a brief report (3-5 sentences) for the hospital administrator explaining your findings in non-technical language

**Your code here:**
```{r exercise-3-2-solution}
# Write your solution here



```

---

## Submission Guidelines

1. Complete all exercises by filling in the code chunks
2. Ensure all code runs without errors
3. Include interpretations and answers to all questions
4. Use appropriate visualizations
5. Write clear, well-commented code

## Evaluation Criteria

Your homework will be evaluated based on:
- Correctness of statistical test selection (25%)
- Proper implementation of tests (25%)
- Quality of visualizations (20%)
- Interpretation of results (20%)
- Code quality and documentation (10%)

---

## Additional Resources

- Review the `continuous_tests.Rmd` file for examples of parametric and non-parametric tests
- Review the `categorical_tests.Rmd` file for examples of Chi-Square and Fisher's tests
- R documentation: `?t.test`, `?wilcox.test`, `?chisq.test`, `?fisher.test`
- Consider effect sizes, confidence intervals, and practical significance
- Always visualize your data before and after analysis

**Good luck!**
