---
title: "Stats 101: Illustrating Key Statistical Tests using the Pima Diabetes Dataset"
author: "Елена Чуклина"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
---

I've selected the **Pima Indians Diabetes Dataset** (a classic in biomedical and machine learning fields) to illustrate these statistical tests. It's a clean, public-domain dataset that offers continuous and binary variables, making it ideal for the required demonstrations.

The dataset contains diagnostic measurements from female Pima Indian patients, 21 years or older, living near Phoenix, Arizona, and includes variables like glucose, blood pressure, BMI, and a diabetes outcome.





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Install and load necessary packages
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(datarium)) install.packages("datarium") 
if(!require(rstatix)) install.packages("rstatix") # For Wilcoxon and Mann-Whitney
library(tidyverse)
library(rstatix) 

# Note: We'll create a synthetic version of the Pima dataset 
# to better illustrate the Paired and ANOVA tests which require specific data structures.
```

## **1. Load and Prepare the Dataset**

We will use the **PimaIndiansDiabetes2** dataset from the `mlbench` package (a clean version of the original UCI dataset).

```{r load_data}
# Install and load mlbench to get the Pima dataset
if(!require(mlbench)) install.packages("mlbench")
library(mlbench)
data("PimaIndiansDiabetes2")
df <- PimaIndiansDiabetes2 %>%
  # Remove NAs for simplicity in this example
  drop_na() %>%
  # Select key variables and rename for clarity
  select(glucose, pressure, pregnant, age, diabetes) %>%
  rename(Outcome = diabetes)

# Inspect the cleaned data structure
print(head(df))
print(skimr::skim(df))
```

-----

## **2. t-Tests (for Means Comparison)**

### **2.1. Independent Samples t-Test**

  * **Goal:** Compare the mean **Glucose** level between patients who tested **Diabetic (pos)** and those who tested **Non-Diabetic (neg)**.
  * **Groups:** Two independent groups (Diabetic vs. Non-Diabetic).
  * **Null Hypothesis ($H_0$):** The mean glucose level is the same for both groups.

<!-- end list -->

```{r independent_ttest}
# Perform the Independent Samples t-test
t_test_ind <- t.test(glucose ~ Outcome, data = df, var.equal = TRUE) # Assuming equal variance for this example

# Print results
cat("### Independent Samples t-Test: Glucose by Diabetes Outcome\n")
print(t_test_ind)

# Visualization
df %>%
  ggplot(aes(x = Outcome, y = glucose, fill = Outcome)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "Glucose Levels by Diabetes Outcome",
       y = "Plasma Glucose Concentration",
       x = "Diabetes Status") +
  theme_minimal()
```

### **2.2. Paired Samples t-Test**

  * **Goal:** To illustrate this, we'll create a **synthetic pre- and post-intervention blood pressure** variable based on the existing `pressure`. We'll simulate a 6-month trial where an experimental diet is hypothesized to lower blood pressure.
  * **Setup:** A single group measured at two time points (paired data).
  * **Null Hypothesis ($H_0$):** The mean difference in pressure (Post - Pre) is zero.

<!-- end list -->

```{r paired_ttest}
# Create a synthetic paired dataset
df_paired <- df %>%
  mutate(
    # 'Pre' is the original blood pressure (baseline)
    pressure_pre = pressure,
    # 'Post' is simulated as 'Pre' - a small, random reduction (due to the "diet")
    pressure_post = pressure_pre - runif(n(), 0, 5) # Simulating a small, random drop
  ) %>%
  select(pressure_pre, pressure_post)

# Perform the Paired Samples t-test
t_test_paired <- t.test(df_paired$pressure_post, df_paired$pressure_pre, paired = TRUE, alternative = "less")

cat("\n### Paired Samples t-Test: Synthetic Blood Pressure (Post vs. Pre-Diet)\n")
print(t_test_paired)
```

### **2.3. One-Sample t-Test**

  * **Goal:** Test if the **mean number of pregnancies** in this Pima Indian cohort is significantly different from a hypothesized **national average of 3** ($\mu_0 = 3$).
  * **Setup:** One sample against a known/hypothesized value.
  * **Null Hypothesis ($H_0$):** The mean number of pregnancies in the sample is 3.

<!-- end list -->

```{r one_sample_ttest}
# Perform the One-Sample t-test against mu = 3
t_test_one <- t.test(df$pregnant, mu = 3)

cat("\n### One-Sample t-Test: Pregnancies vs. National Average (mu = 3)\n")
print(t_test_one)
```

-----

## **3. Analysis of Variance (ANOVA)**

### **3.1. One-Way ANOVA**

  * **Goal:** Compare the mean **Glucose** level across different **Age Groups**.
  * **Setup:** Three or more independent groups (requires grouping a continuous variable).
  * **Null Hypothesis ($H_0$):** The mean glucose is the same across all age groups.

<!-- end list -->

```{r anova}
# 1. Create age groups (a factor with 3+ levels)
df_anova <- df %>%
  mutate(
    age_group = case_when(
      age < 30 ~ "Young Adult",
      age >= 30 & age < 45 ~ "Middle Adult",
      TRUE ~ "Older Adult"
    ) %>%
    as.factor()
  )

# 2. Check group counts to ensure adequate size
print(table(df_anova$age_group))

# 3. Perform One-Way ANOVA
anova_model <- aov(glucose ~ age_group, data = df_anova)

cat("\n### One-Way ANOVA: Glucose by Age Group\n")
print(summary(anova_model))

# Visualization
df_anova %>%
  ggplot(aes(x = age_group, y = glucose, fill = age_group)) +
  geom_boxplot() +
  labs(title = "Glucose Levels by Age Group",
       y = "Plasma Glucose Concentration",
       x = "Age Group") +
  theme_minimal()
```

-----

## **4. Non-Parametric Tests (Rank-Based)**

We often use non-parametric tests when the assumption of **normality is violated** or the data is **ordinal/ranked**.

### **4.1. Mann-Whitney U Test (Wilcoxon Rank Sum)**

  * **Non-parametric equivalent of:** Independent Samples t-Test.
  * **Goal:** Compare the **distributions (medians)** of **Body Mass Index (BMI)** between Diabetic and Non-Diabetic patients.
  * **Setup:** Two independent groups.


```{r mann-whitney}
# Assuming BMI is not normally distributed (we would typically test this)
test_mw <- wilcox.test(bmi ~ Outcome, data = df, exact = FALSE)

cat("\n### Mann-Whitney U Test: BMI by Diabetes Outcome\n")
print(test_mw)
```

### **4.2. Wilcoxon Signed-Rank Test**

  * **Non-parametric equivalent of:** Paired Samples t-Test.
  * **Goal:** Compare the **distributions (medians)** of the **synthetic paired blood pressure data** created earlier.
  * **Setup:** Two related (paired) samples.


```{r wilcoxon_signed_rank}
# The 'df_paired' data has 'pressure_pre' and 'pressure_post'
test_wsr <- wilcox.test(df_paired$pressure_post, df_paired$pressure_pre, paired = TRUE, alternative = "less")

cat("\n### Wilcoxon Signed-Rank Test: Synthetic Paired Blood Pressure (Post vs. Pre-Diet)\n")
print(test_wsr)
```

```{r}
# Perform the Independent Samples t-test
t_test_ind <- t.test(glucose ~ Outcome, data = df, var.equal = TRUE) # Assuming equal variance for this example

# Print results
cat("### Independent Samples t-Test: Glucose by Diabetes Outcome\n")
print(t_test_ind)

# Visualization
df %>%
  ggplot(aes(x = Outcome, y = glucose, fill = Outcome)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "Glucose Levels by Diabetes Outcome",
       y = "Plasma Glucose Concentration",
       x = "Diabetes Status") +
  theme_minimal()
```


```{r}
# Create a synthetic paired dataset
df_paired <- df %>%
  mutate(
    # 'Pre' is the original blood pressure (baseline)
    pressure_pre = pressure,
    # 'Post' is simulated as 'Pre' - a small, random reduction (due to the "diet")
    pressure_post = pressure_pre - runif(n(), 0, 5) # Simulating a small, random drop
  ) %>%
  select(pressure_pre, pressure_post)

# Perform the Paired Samples t-test
t_test_paired <- t.test(df_paired$pressure_post, df_paired$pressure_pre, paired = TRUE, alternative = "less")

cat("\n### Paired Samples t-Test: Synthetic Blood Pressure (Post vs. Pre-Diet)\n")
print(t_test_paired)
```


